---
title: 'Heteroskedasticity'
date: "2023/6/22"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


> Learning points:
>
> * calculate heteroskedasticity robust standard errors using `coeftest()` function
> * F-test under heteroskedasticity assumption using `linearHypothesis()` function
> * use `ggplot()` function of the `ggplot2` package

This document is adapted from the article [How to Perform Weighted Least Squares Regression in R](https://www.statology.org/weighted-least-squares-in-r/) on statology.org.



# Create the Data

The following code creates a data frame that contains the number of hours studied and the corresponding exam score for 16 students:

```{r}
df <- data.frame(hours=c(1,1,2,2,2,3,4,4,4,5,5,5,6,6,7,8),
                 score=c(48,78,72,70,66,92,93,75,75,80,95,97,90,96,99,99))
```

# Perform Linear Regression

```{r}
# fit simple linear regression model
model <- lm(score ~ hours, data=df)
summary(model)
```

# Test for Heteroscedasticity

Next, we’ll create a residual vs. fitted values plot to visually check for heteroscedasticity:

```{r}
# create residual vs. fitted plot
plot(fitted(model), resid(model), xlab="Fitted Values", ylab="Residuals")
abline(0,0)        # add a horizontal line at 0
```

We can see from the plot that the residuals exhibit a “cone” shape – they’re not distributed with equal variance throughout the plot. 

To formally test for heteroscedasticity, we can perform a Breusch-Pagan test:

```{r message=FALSE}
# perform Breusch-Pagan test
library(lmtest)     # load the required package
bptest(model)       # perform Breusch-Pagan test
```

The Breusch-Pagan test uses the following null and alternative hypotheses:

* Null Hypothesis ($H_0$): Homoscedasticity is present (the residuals are distributed with equal variance)
* Alternative Hypothesis ($H_a$): Heteroscedasticity is present (the residuals are not distributed with equal variance)

Since the p-value from the test is $0.0466$ we will reject the null hypothesis and conclude that heteroscedasticity is a problem in this model.

# Perform Weighted Least Squares Regression

Since heteroscedasticity is present, we will perform weighted least squares by defining the weights in such a way that the observations with lower variance are given more weight:


