---
title: 'Activity 2 - MRM: Estimation'
author: "AW"
date: "2023/6/22"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Running OLS regression

> Learning points:
> 
> * library `foreign`


```{r}
# Load the required package
library(foreign)
```

```{r}
# Read the data
data <- read.dta("Activity2/CARD.dta")
```

The data has 34 variables and 3010 observations. We only care about a few of these, namely "wage", "educ", "IQ", "black", and "exper".

```{r}
data_for_analysis <- subset(data, select = c(wage, educ, IQ, black, exper)) 
head(data_for_analysis)
```

```{r}
# Create log wages
data_for_analysis$lwage <- log(data_for_analysis$wage)
```

```{r}
# Run OLS regression
bivariate_OLS <- lm(lwage ~ educ, data=data_for_analysis)
summary(bivariate_OLS)
```

```{r}
# plot residuals against education
residual <- residuals(bivariate_OLS)    # get the residuals
plot(data_for_analysis$educ, residual,
     xlab = "Years of education", ylab = "Residuals",
     main = "Residuals vs. Years of education")
```

```{r}
# get R squared
summary(bivariate_OLS)$r.squared
```

```{r}
# get adjusted R squared
summary(bivariate_OLS)$adj.r.squared
```

```{r}
# run multiple OLS regression
multivar_OLS <- lm(lwage ~ educ + exper, data=data_for_analysis)
summary(multivar_OLS)
```

```{r}
# running auxilliary regression
aux_reg <- lm(exper ~ educ, data=data_for_analysis)
summary(aux_reg)
```

Adding a second control: IQ

```{r}
# IQ has missing value
summary(data_for_analysis$IQ)
```

We include the option of `na.action = na.exclude` in the `lm()` function to exclude the missing value. In fact, `R` automatically drops all missing observations in a regression. However, it is good practice to include option as it will force you to think about what observations you drop from your sample an possible selection issues.

```{r}
# runing second multiple OLS regression
multivariate_OLS <- lm(lwage ~ educ + IQ + exper, data=data_for_analysis,
                       na.action=na.exclude)
summary(multivariate_OLS)
```

