---
title: 'Simulation-Repeated'
date: "2023/6/22"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This file is based on Script 1.38 and 1.39 of Heiss' book.

```{r}
# set the random seed
set.seed(123456)

#initialize ybar to a vector of length r=10000 for storing results
ybar <- numeric(10000)

# repeat 10000 times
for (j in 1:10000) {
  # Draw a sample and store the sample mean in ybar
  sample <- rnorm(100, 10, 2)
  ybar[j] <- mean(sample)
}

# the first 20 of 10000 estimates
ybar[1:20]

# simulated mean
mean(ybar)

# simulated variance
var(ybar)

# simulated density
plot(density(ybar))
curve( dnorm(x, 10, sqrt(var(ybar))), add = TRUE, lty = 2)
```

# Line types in R : `lty`

The different line types available in R are shown in the figure hereafter. The argument `lty` can be used to specify the line type. To change line width, the argument `lwd` can be used.

line type (`lty`) can be specified using either text (“blank”, “solid”, “dashed”, “dotted”, “dotdash”, “longdash”, “twodash”) or number (0, 1, 2, 3, 4, 5, 6). Note that `lty` = “solid” is identical to `lty=1`.

```{r}
# Solid line (by default)
plot(1:10, 1:10, type="l")
# Use dashed line type
plot(1:10, 1:10, type="l", lty=2)
# Change line width
plot(1:10, 1:10, type="l", lty=2, lwd=3)
```

Note that the option `type="l"` ("l" is the small letter of "L", not "1" (one))means it is a line plot. If this option is skipped, the default plot is a dot plot.

```{r}
#Line types
#++++++++++++++++++++++++++++++++++++++++++++
generateRLineTypes<-function(){
  oldPar<-par()
  par(font=2, mar=c(0,0,0,0))
  plot(1, pch="", ylim=c(0,6), xlim=c(0,0.7),  axes=FALSE,xlab="", ylab="")
  for(i in 0:6) lines(c(0.3,0.7), c(i,i), lty=i, lwd=3)
  text(rep(0.1,6), 0:6, labels=c("0.'blank'", "1.'solid'", "2.'dashed'", "3.'dotted'",
                                 "4.'dotdash'", "5.'longdash'", "6.'twodash'"))
  par(mar=oldPar$mar,font=oldPar$font )
}
generateRLineTypes()
```

See the [online reference](http://www.sthda.com/english/wiki/line-types-in-r-lty#:~:text=line%20type%20(lty)%20can%20be,is%20identical%20to%20lty%3D1.) for the above discussion.


# Normal curve that describes the histogram

```{r}
# other than the "unit Normal"
w<-rnorm(10000, mean=10, sd=2) 
hist(w, col="red", freq=F, xlim=10+c(-5,5), ylim = c(0,0.2))
curve( dnorm(x, mean=10,sd=2), 5, 15, add=T, col="blue")
```

```{r}
# unit normal
w<-rnorm(1000) 
hist(w,col="red",freq=F,xlim=c(-5,5))
curve(dnorm,-5,5,add=T,col="blue")
```